Fluent statistical computing interfaces for biological data analysis

### 2020-04-15 Rmd/01-intro.Rmd
add citation

This thesis focuses on three core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation (in particular we 
focus on the exploration of genomics and transcriptomics data).
It is organised in the same manner that you would iterate back and
forth between these tasks  during an analysis. To begin you wrangle your data into a structure 
that you can understand and use; section \@ref(sec:tidygranges) introudces a
grammar-based framework for transforming genomics data that is described in 
chapter \@ref(ch:plyranges). Next, you integrate data and model outputs
over genomic regions to gain biological insight; \@ref(sec:ir) introduces a
framework for incorporating and discovering 'interesting' genomic regions,
described further in chapters \@ref(ch:fluentGenomics) and \@ref(ch:intron).
Third, you cycle between model-based representations and visualisations
of your data; section {#sec:va} introduces a visual analytics approach for
understanding non-linear dimension reduction techniques and is described in
detail in chapter \@ref(ch:tsne). Lastly, you take a step back and look at
the implications and limitations of your work; in chapter \@ref(ch:conclusion)
I describe the impacts of my work and plans for future research.

@@

- talking about chapters 3 and 4 here
- having the grammar doesn't impair interoperability, in fact we show it
can be used to glue two different approaches
- the long form representation can be taken even further... 
- recasting EDA problems as search, we want to find regions measured along the
genome that display a characteristic of interest, and then link that to raw 
data plots.

## Visual analytics for dimension reduction workflows

>>>>

This thesis focuses on three core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation (in particular we 
focus on the exploration of genomics and transcriptomics data).
It is organised in the same manner that you would iterate back and
forth between these tasks  during an analysis. To begin you wrangle your data 
into a structure that you can understand and use; section \@ref(sec:tidygranges)
introudces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, you integrate data and model 
outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, you cycle between model-based representations and 
visualisations of your data; section {#sec:va} introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques and is described in
detail in chapter \@ref(ch:tsne). Lastly, you take a step back and look at
the implications and limitations of your work; in chapter \@ref(ch:conclusion)
I describe the outpus of my thesis and plans for future research.

@@

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study. While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
we describe a simpler end-to-end workflow for integrating 
results along the genome using *plyranges*. This workflow also shows that
our grammar based approach does not impair interoperability between
the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
together.

Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
data by developing a new software tool called *superintronic*. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use *superintronic* to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data.

## Visual analytics for dimension reduction workflows {#sec:va}
